sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-01-14 doc/appendix-pcod-sdmTMB.Rmd
experiment with different example headings; minor edits to owl section and both appendices

We will fit a geostatistical spatiotemporal model with sdmTMB for the purposes of index standardization. We will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound. The density units are kg/km^2^. Here, X and Y are coordinates in UTM zone 9.

>>>>

We will fit a geostatistical spatiotemporal model with sdmTMB for the purposes of index standardization. We will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound. The density units are kg/km^2^. Here, X and Y are UTM zone 9 coordinates in kilometres. 
If our spatial coordinates were not already in an equidistant projection, we would first transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region.

### 2022-01-14 doc/appendix-snow-sdmTMB.Rmd
experiment with different example headings; minor edits to owl section and both appendices

We can then compress the annual predictions into mean values for plotting the spatially varying effect of NAO.
Most simply, we can extract the overall coefficients using the `b <- tidy(m)` and add them together with the spatially varying coefficients `zeta_s`.
However, this method does not give us spatially explicit estimates of uncertainty.
To do this, we can use the output from the joint precision matrix simulation draws by using the `predict()` function again but this time indicating a number of simulation draws (e.g., `sims = 500`) and what component we want estimated (`sims_var = "zeta_s"`). 
We also extract the same number of drams for (this time `n_sims = 500`) in the `spread_sims()` function for the main effect of NAO. 
The simulation draws for both should then be combined before adding the median and lower and upper quantiles from these draws to our prediction data frame.

>>>>

We can then compress the annual predictions into mean values for plotting the spatially varying effect of NAO.
Most simply, we can extract the overall coefficients using the `b <- tidy(m)` and add them together with the spatially varying coefficients `zeta_s`.
However, this method does not give us spatially explicit estimates of uncertainty.
To do this, we can get the output from the joint precision matrix simulation draws by using the `predict()` function again, but this time indicating a number of simulation draws (e.g., `sims = 500`) and what component we want estimated (`sims_var = "zeta_s"`). 
We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function (this time `n_sims = 500`). 
The simulation draws for both can then be combined before calculating the median and lower and upper quantiles from these draws and adding these values to our prediction data frame.

### 2022-01-14 doc/paper.Rmd
experiment with different example headings; minor edits to owl section and both appendices

# Introductory example with fish survey biomass

An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
<!-- TODO: Since this is not particular to your package, maybe the e.g. for how to do this should be moved to the appendix? I do explain the transformation for the owl data, but it could be added to the pcod appendix too? -->
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.

@@

(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
(A) SPDE mesh (lines) combined with the trawl survey observations (points).
Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
The circle area corresponds to density of Pacific cod for individual trawls.
(B) Spatial random field:
these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
(D) Conditional effect of depth modelled as a penalized smoother. These predictions are made omitting the spatial random field.
(E) Standardized area-weighted population index derived from a spatiotemporal version of this model. The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
<!-- LB: I wasn't sure which panel the statement about omitting the spatial random field applied to, but moved things around. -->

@@

# Spatially varying coefficient example with citizen science data

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid latitudes of North America in much greater numbers in some winters than others.
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (Appendix 3).

@@

We found a weak average positive effect of annual mean NAO on overall counts, but a marked southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao)).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.

>>>>

<!-- # Introductory example with fish survey biomass -->
# Introductory example using fish survey data

An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
<!-- TODO: Since this is not particular to your package, maybe the e.g. for how to do this should be moved to the appendix? I do explain the transformation for the owl data and added a sentence to the pcod appendix, but maybe still important to include here as well? -->
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.

@@

(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
(A) SPDE mesh (lines) combined with the trawl survey observations (points).
Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
The circle area corresponds to biomass density of Pacific cod caught on individual trawls.
(B) Spatial random field:
these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
(D) Conditional effect of depth modelled as a penalized smoother. These predictions are made omitting the spatial random field.
(E) Standardized area-weighted population index derived from a spatiotemporal version of this model. The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
<!-- LB: I wasn't sure which panel the statement about omitting the spatial random field applied to, but moved things around. -->

@@

<!-- # Spatially varying coefficient example with citizen science data -->
# 2. Estimate a spatially varying effect from citizen science data

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid latitudes of North America in much greater numbers in some winters than others.
The reasons for this interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.

@@

We found a weak average positive effect of annual mean NAO on overall counts, but a marked southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Appendix 3).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
